---
title: "Creating State Maps"
author: "Rachel Carroll, Mazama Science"
date: "July 25, 2020"
output: rmarkdown::html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Objective

The goal of this document is to introduce the `stateMap()` function in the 
**MazamaSpatialPlots** package. This document will serve as a tutorial for 
collecting data and creating customizable maps of U.S. states and territories. 

## Getting State-level Data

The `stateMap()` function requires two forms of state-level data. First is a 
Spatial Polygons Data Frame (SPDF) containing state level polygons. The SPDF must 
include the variable `stateCode` in it's data `@slot`. The default SPDF is 
"USCensusStates_02", which is contained in the package data. Higher or lower 
resolution US Census state SPDFs can downloaded from
http://data.mazamascience.com/MazamaSpatialUtils/Spatial_0.7/
or generated and loaded using 
`MazamaSpatialUtils::convertUSCensusStates()`. 

The second required form of data is a regular dataframe that contains the 
variable `stateCode` as well as a variable of interest. The variable of interest 
from this dataset is indicated using the `parameter` input. It is used to color 
states in the generated chloropleth map.   

Below, there are three examples of obtaining, summarizing, and mapping 
state-level data. The first two use example dataframes from the package and the 
last uses `MazamaCoreUtils::html_getTable()` to easily pull data from a website.

### Using package dataframe `example_US_stateObesity`

In this example, the package-internal dataframe, `example_US_stateObesity`, 
is used directly:

```{r plot_obesity, warning = FALSE, message = FALSE }
library(MazamaSpatialPlots)

stateMap(
  data = example_US_stateObesity,
  parameter = 'obesityRate',
  title = "Obesity Rate in U.S. States"
)
```

### Using package dataframe `example_US_countyCovid`

In this example, the county-level package dataframe, `example_US_countyCovid`, 
is aggregated by state and then used:

```{r plot_covid, warning = FALSE, message = FALSE }
library(dplyr)

example_US_stateCovid <- 
  example_US_countyCovid %>%
  dplyr::group_by(stateCode) %>%
  dplyr::summarise(stateCases = sum(cases), stateDeaths = sum(deaths)) %>%
  dplyr::mutate(stateDeathRate = 100*stateDeaths/stateCases ) %>%
  dplyr::select(c("stateCode", "stateCases", "stateDeaths", "stateDeathRate"))

stateMap(
  data = example_US_stateCovid,
  parameter = 'stateDeathRate', 
  title =  "COVID Fatality Rates (%) on June 12, 2020"
)
```

### Using data found online
State-level data of interest can be found online and easily scraped using 
`MazamaCoreUtils::html_getTable()` to parse all table elements from a website. 
Using `html_getTable()` in conjunction with `stateMap()` makes it very easy to 
extract and visualize data from the internet.

```{r html_getTable_and_stateMap, warning = FALSE, message = FALSE }

URL <- "https://www.patriotsoftware.com/blog/accounting/average-cost-living-by-state/"
livingCostData <- MazamaCoreUtils::html_getTable(URL)

livingCostData <- 
  livingCostData %>%
  dplyr::mutate(
    stateCode = MazamaSpatialUtils::US_stateNameToCode(.data$"State"),
    avgAnnualWage = as.numeric(gsub('[$,]', '', .data$"Annual Mean Wage (All Occupations)")),
    avgMonthlyRent = as.numeric(gsub('[$,]', '', .data$"Median Monthly Rent")),
    rentWagePercent = 100*12*avgMonthlyRent/avgAnnualWage,
    .keep = "none"
    ) 

stateMap(
  data = livingCostData,
  parameter = 'rentWagePercent',
  title = "Average Percentage of Income Spent on Rent"
)
```


## Customizing with Function Parameters

In the section above examples, the `stateMap()` inputs `data`, `parameter`, 
`state_SPDF`, and `title` are used to create maps. This section will demonstrate 
how to use the other input parameters to customize your map. 

### Using palette, breaks, and stateBorderColor

The `palette`, `breaks`, and `stateBorderColor` parameters dictate the coloring 
of your map. The shade(s) are defined with `palette` and the distribution of 
color across the map is defined with `breaks`. As expected, `stateBorderColor` 
defines the state border color. 

To make the most of these parameters, see the following references for R colors 
and palettes:

  * <http://sape.inf.usi.ch/quick-reference/ggplot2/colour>
  
  * <https://www.r-graph-gallery.com/38-rcolorbrewers-palettes.html>

In this example, `breaks` is used to create a coarser coloring scheme and 
`palette` is used to customize the exact color for each obesity rate level.

```{r plotColors, warning = FALSE, message = FALSE }

stateMap(
  data = example_US_stateObesity,
  parameter = 'obesityRate',
  palette = c("lightblue", "orange", "red"),
  breaks = c(20, 27, 34, 40),
  stateBorderColor = "black",
  title = "Obesity Rate in U.S. States"
)

```

In this example, `breaks` is used to create a more detailed coloring scheme and 
the blue to purple color palette is chosen.

```{r plotStates, warning = FALSE, message = FALSE }

stateMap(
  data = example_US_stateObesity,
  parameter = 'obesityRate',
  palette = 'BuPu',
  breaks = seq(20, 38, 3),
  stateBorderColor = 'white',
  title = "Obesity Rate in U.S. States"
)

```


### Using conusOnly and stateCode

The `conusOnly` and `stateCode` parameters define which states will be included 
in the map. If `stateCode` is defined, then `conusOnly" will be ignored. If 
`stateCode` is not defined, then `conusOnly` specifies if only the continental 
US will be displayed. When `conusOnly = FALSE`, the continental U.S., Alaska, 
Hawaii, and U.S. Territories will be included. 

This example builds upon the previous example and includes `stateCode` 
specification to create a map of Western states. 

```{r use_stateCode, warning = FALSE, message = FALSE }

stateMap(
  data = example_US_stateObesity,
  parameter = 'obesityRate',
  palette = 'PuBuGn',
  breaks = seq(23, 31, 2),
  stateCode = c("CA", "NV", "OR", "WA", "ID"),
  stateBorderColor = 'black',
  title = "Obesity Rate in Western U.S."
)


```

The following example uses `conusOnly = FALSE`.

```{r use_conusOnly, warning = FALSE, message = FALSE }

stateMap(
  data = example_US_stateObesity,
  parameter = 'obesityRate',
  palette = 'YlOrRd',
  breaks = seq(20, 38, 3),
  conusOnly = FALSE,
  stateBorderColor = 'black',
  title = "Obesity Rate in U.S."
)


```


### Using projections

In general, the default projections in `stateMap()` are appropriate for common 
state groupings. However, the projection parameter can be used to override the 
default projections as needed. 

Below is an unusual group of states with the default projection.

```{r defaultprojection, warning = FALSE, message = FALSE }
northernStates <- c("AK", "WA", "ID", "MT", "ND", "MN", "WI", 
                    "MI", "OH", "PA", "NY", "VT", "NH", "ME")

stateMap(
  data = example_US_stateObesity,
  parameter = 'obesityRate',
  palette = 'YlOrRd',
  breaks = seq(23, 33, 2),
  stateCode = northernStates,
  stateBorderColor = 'black',
  title = "Obesity Rate in Northern U.S. States"
)

```

The projection is oddly rotated in this case. To remedy this, a manual projection 
can be defined: 

```{r manualprojection, warning = FALSE, message = FALSE }

myproj <- "+proj=lcc +lat_1=32.5 +lat_2=71.4 +lat_0=51.9 +lon_0=-102.9 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs"
stateMap(
  data = example_US_stateObesity,
  parameter = 'obesityRate',
  palette = 'YlOrRd',
  breaks = seq(23, 33, 2),
  stateCode = northernStates,
  projection = myproj,
  stateBorderColor = 'black',
  title = "Obesity Rate in Northern U.S. States"
)

```


## Customizing with the **tmap** Package

One benefit of using `stateMap()` is that the output is a _ggplot2_ object. 
Therefore, maps can be further customized using functionality from the **tmap**
and **ggplot2** packages. Many visual adjustments including legend and title 
locations, fonts, and background colors can be made by simply appending your 
`stateMap()` plot with arguments in `tmap::tm_layout()`. Furthermore,
`tmap::tm_style()` can be used to leverage built in styles and 
`tmap::tm_compass()` can be used to add a compass to your map. The three 
examples below demonstrate how to build customized visualizations using both 
`stateMap()` parameters and **tmap** functionality.

### Using tmap::tm_layout()

```{r tmapStyling1, warning = FALSE, warning = FALSE, message = FALSE }

stateMap(
  data = example_US_stateObesity, 
  parameter = "obesityRate",
  breaks = seq(20,38,3), #increasing color detail
  conusOnly = FALSE ,
  stateBorderColor = 'black',
) +
  tmap::tm_layout(
    frame = TRUE,
    main.title = 'Obesity Rates in U.S. States and Territories',
    main.title.position = c("center", "top"),
    title.fontface = 2,
    fontfamily = "serif",
    bg.color = "grey85", 
    inner.margins  = .05,
    legend.position = c('right', 'top')
  ) 

```

### Using tmap style "classic" and a compass

```{r tmapStyling2, warning = FALSE, message = FALSE }

stateMap(
  data = example_US_stateObesity, 
  parameter = "obesityRate", 
  breaks = seq(20,40,4),
) +
  tmap::tm_style(
    'classic'
  ) +
  tmap::tm_layout(
    title = 'Obesity Rate by State',
    title.position = c("center", "top"),
    title.size = 1.2,
    inner.margins  = .08
  ) +
  tmap::tm_compass() 
  

```

### Polishing  New England regional map

```{r tmapStyling3, warning = FALSE, message = FALSE }
  
stateMap(
 data = example_US_stateObesity,
 parameter = "obesityRate",
 stateCode = c('ME', 'NH', 'VT', 'MA', 'RI', 'CT'),
 stateBorderColor = 'black',
 title = 'Obesity Rates in New England'
) +
 tmap::tm_layout(
   frame = TRUE,
   frame.double.line = TRUE,
   title.size = 1.2,
   title.fontface = 2,
   fontfamily = "serif",
   bg.color = "grey85",
   inner.margins  = .08
 )

```

## Conclusion

The `stateMap()` function allows us to create attractive maps very efficiently. 
When used alongside `MazamaCoreUtils::html_getTable()`, U.S. state data can be 
procured and visualized in very few lines of code. Additionally, the combination 
of these two functions provides a great deal of flexibility. The `html_getTable()` 
function opens up an endless source of data from the internet while `stateMap()` 
can create highly customized visualizations through direct inputs and by 
harnessing the functionality of the **tmap** package. 






